Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-08-06 Report/Sections/data.tex
Update Report

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale and the distribution of acceleration within each behavioral state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range are ignored as well. A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data are smoothed by taking a moving average within a window of 1/10th of a second. Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, $Y$ is a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. The fine-scale process $Y^*$ is the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.

\subsection{Model Selection}

A lag plot is used to determine where auto-correlation should be incorporated into the model (see Section S-2.1). The coarse-scale level exhibits no significant amount of auto-correlation and no obvious intricate structure prior to fitting the model. As such, we select a simple HMM to model the coarse-scale process. The dive durations $Y_t$ are assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics. A window size of two-seconds is selected since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data. The acceleration data was also collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than in Equation (\ref{eqn:z}):
%
\begin{equation*}
    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right), \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation*}
%
Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar. We calculate $Z_t^{*(2)}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1). Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$. In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$. In words, each axis of acceleration has its own mean and variance, but all axes share an auto-correlation parameter.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$. Therefore, a simple HMM is used to model $Z^{*(2)}$. Given that $X^*_{t,t^*} = i^*$, the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%

@@

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM. We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that dive types 1 and 2 share subdive states. In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
Recall that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed so far. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT. In addition to the CarHHMM-DFT we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM. However, the Fourier sums $Z^{*(2)}$ are omitted from this model altogether. The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have emission distributions identical to the CarHHMM-DFT when applicable. Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.

>>>>

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. HMMs have been used to understand the how diving behaviors of various species are affected by disturbances (e.g. https://projecteuclid.org/euclid.aoas/1491616885  and https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2044). For NRKWs, we are interested in categorizing different diving behaviors, with a particular focus on differentiating potential foraging dives. NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale and the distribution of acceleration within each behavioral state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using tag [NAME OF TAG - ask Sarah]. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25pm and 2:37pm and between 4:07 and 5:07, when the tagging technology malfunctioned. A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data are smoothed by taking a moving average within a window of 1/10th of a second. Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the observed coarse-scale observation set $y$ is a time series of dive durations in seconds ($s$) with corresponding unobserved dive types $x$. The fine-scale observations $y^*$ are the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $x^*$.

\subsection{Model Development and Selection}

To choose how to model the coarse-scale diving data, we first inspect lag plots to determine whether auto-correlation should be incorporated (see Section S-2.1). The coarse-scale level exhibits no significant amount of auto-correlation and no obvious intricate structure prior to fitting the model. As such, we select a simple HMM to model the coarse-scale process. The dive durations $Y_t$ are assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

In contrast, the fine-scale acceleration data exhibits significant sinusoidal behavior. Thus, we transform $Y^*$ to $Z^*$ using the DFT for this model. We select a window size of two-seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data. The acceleration data was collected in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than in Equation (\ref{eqn:z}):
%
\begin{equation*}
    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \qquad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation*}
%
Each observation is made up of $\Zone_{t,t^*}$ (a 3-dimensional vector of average acceleration) and $\Ztwo_{t,t^*}$ (a scalar describing ``wiggliness"). We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (Section S-2.1). Therefore, a CarHMM is used to model $\Zone_{t,t^*}$. In particular, if $X^*_{t,t^*} = i^*$, then we assume $\Zone_{t,t^*}$ is Normally distributed with unknown parameters $\phi_A^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(i^*)} \in \mathbb{R}^3$. In words, each axis of acceleration has its own mean and variance, but all axes share an auto-correlation parameter.
%
%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
%

While $\Ztwo$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $\Zone$, and the biological interpretation of auto-correlation within $\Ztwo$ is less clear. In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$. Therefore, a simple HMM is used to model $\Ztwo$. Given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_W^{*(i^*)}$ and variance $\sigma_W^{*(i^*)}$.
%
%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
%

@@

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $\Zone$ is modelled with a CarHMM, and $\Ztwo$ is modeled with an HMM. We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that dive types 1 and 2 share subdive states. In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)}\},\{\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT. In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $\Zone$ \textit{and} $\Ztwo$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $\Zone$ using a CarHMM. However, the Fourier sums $\Ztwo$ are omitted from this model altogether. The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\Zone$ with a CarHMM and $\Ztwo$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}

Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $\Ztwo$, and the CarHMM-DFT lacks a hierarchical structure.

### 2020-08-06 Report/Sections/discussion.tex
Update Report

In this work, we deal with both temporal dependence between curves and a discrete number of curve types by modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves. Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation between curves. One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions. The parameters of this fine-scale model then in turn depend upon the curve type. Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.

@@

To deal with these problematic fine-scale processes, we detail a moving-window transformation over the fine-scale observations. For periodic behaviour in particular, we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale. On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it. After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT. This model effectively decodes both dive types and subdive states of a killer whale while also taking into account the interaction between these two levels.

This solves an important problem, as the energetic requirements of killer whales together with their energetic intake from predation is directly related to their survival \citep{Williams:2009,Noren:2011}. Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}. Accelerometers are an appealing alternative, but require decoding both the behavioral state \citep{Dot:2016} and the dive type \citep{Hastie:2006} of the animal in order to effectively estimate energy expenditure. The CarHHMM-DFT solves this problem by decoding both behaviors simultaneously at different scales. It also describes the distribution of acceleration within each hidden state. All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.

While we combine the HMM models of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models. For example, HMMs require observations to be taken at regular intervals. Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue, but we do not describe them in detail here. This is because most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit. In addition, we show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions. This implies that a CarHMM is interpretable like a continuous-time model, but as easy to fit as a traditional HMM.

Another consideration when using HMMs to build complex models is the importance of effectively fit the data in question while avoiding \textit{over}-fitting or computationally expensive parameter estimation. One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states. Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states. However, this model is not sufficient if ecologists wish to understand to joint relationship between the dive type and intra-dive behaviour.

@@

Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models. When model complexity is kept under control, a hierarchical model constructed using this processes can be both highly flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale. It is our hope that researchers adapt this methodology to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.

>>>>

In this work, we deal with both temporal dependence between curves and a discrete number of curve types by using either an HMM or CarHMM, depending upon the level auto-correlation between curves. To help select the appropriate model, lag plots can be used to test if there is significant auto-correlation between curves. One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as the HMM emissions. The parameters of this fine-scale model then in turn depend upon the curve type. Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.

@@

To deal with these problematic fine-scale processes, we detail a moving-window transformation over the fine-scale observations. For periodic behaviour in particular, we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $\Ztwo$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale. [MAYBE ADD SOMETHING ABOUT THE CONFIRMATION OF THIS IN THE SIMULATION STUDY.] For simpler transformations that do not directly account for periodicity but for which me may expect temporal correlation (e.g., moving average), we suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it. Our simulation study showed how ignoring this auto-correlation can negatively bias standard error estimates.

Accelerometers and time depth recorders are powerful tools to understand the diving behavior and energetic requirements of NRKWs, information essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}. But this biologging data require decoding the behavioral state \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand estimate energy expenditure. The CarHHMM-DFT model we developed solves this problem by characterising dives into different types. It also describes the distribution of acceleration within each hidden state, providing a deeper understanding of the tri-axial movement, and thus fine-scale energy expenditure, of the animal. [ADD a ctation that linked up tri axial movement to fluking and energy expenditure. Check maybe some of the papers Courtney sent.] [Also add a sentence that this model is likely applicable to the many diving animals such as ..., maybe cite Adams, Vianey or anything that has both dive and accelerometer data.]

While we combine the HMMs of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models. For example, HMMs require observations to be taken at regular intervals. Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue. However, most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit. We show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions. This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.

When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding \textit{over}-fitting or computationally expensive parameter estimation. One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states. Regardless, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states. However, this model is not sufficient if ecologists wish to understand the joint relationship between the dive type and intra-dive behaviour.

@@

Our work provides a flexible framework to model functional time-series data using HMMs. When model complexity is kept under control, a hierarchical model constructed using the main building blocks proposed can be both highly flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a structure using an ecological example, where we used a complex HMM to classify the coarse and fine scale diving behavior of a killer whale. Such model will be applicable to similar datasets. In addition, as temporally dependent state-switching processes are common (IF YOU HAD CITATIONS HERE YOU I'D be AMZING-not necessary, but would seal the deal.), we believe that researchers can adapt this methodology to perform inference on a wide range of time-series data in other fields.

### 2020-08-06 Report/Sections/intro.tex
Update Report

Tracking technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by tracking devices that collects biologging data (CITATION). Tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. These data provide biologists with a wealth of knowledge about animal behaviour, but parsing this vast amount of information is a challenge for statisticians.

@@

Movement data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}). In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time. The dive profile of marine mammals can be viewed as a curve, and Fu (CITATION) showed that [methods used by Eric and Nancy] is useful to classify [blah blah blah]. However, now many such animals are tracked with multiple devices providing information at different scales. For example, in our case study, a single dive of a killer whale (\textit{Orcinus orca}) can be viewed as having several curves where acceleration and depth are functions of time.

FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) (CITATION would be good - Fu's discussion? Our Vianey's paper or Timo Adam's paper?). Our case study data clearly demonstrate this common additional dependencies between curves. The dive profile (Figure \ref{fig:data}A) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing) clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time. Previous studies have categorized these dives into different types, each associated with a different behavior, including resting, traveling, or foraging \citep{Tennessen:2019b}.
Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. The combination of apparent behavioral changes combined with periodic behavior makes this accelerometer data difficult to model using simple stochastic processes. [Here if you can say something about how common this combined periodicity and behavioral dependece is in other accelerometer data, e.g. in human with a citation, you'd really effectively broaden the generality of the problem.]

>>>>

Tracking technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by tracking devices that collect biologging data \citep{Borger:2020}. Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of knowledge about human and animal behaviour, but parsing this vast amount of information is a challenge for statisticians.

@@

Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}). In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time. The dive profile of a marine mammal can also be viewed as a curve, and \citet{Fu:2017} showed that estimating the amplitude and phase variation of elephant seal curves is useful to classify dive types, although they ignore sequential dependence between dives. 
%However, now many such animals are tracked with high-frequency tags over a period of several dives, providing information at different time scales. 
In our case study, the dive profile of a killer whale (\textit{Orcinus orca}) can be viewed as a sequence of curves where three orthogonal axes of acceleration are each functions of time.

FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}. Our case study data clearly demonstrate this common additional dependencies between curves. The dive profile (Figure \ref{fig:data}A) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing) clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time. Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour. These behaviours can include resting, travelling, and foraging \citep{Tennessen:2019b}.
Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. The combination of apparent behavioural changes combined with periodic behaviour makes this accelerometer data difficult to model using simple stochastic processes. Another example of behaviourally-dependent periodicity is in the field of machine health, where \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks. The raw acceleration data exhibit extreme periodicity (vibration) that depends upon the behaviour of the train as the load and speed of the train change.
